---
output:
  word_document: 
    toc: true
  html_document: default
---
# Week 3 Homework - Applying Good Design Principles using mtcars

## Load the dataset
```{r}

data("mtcars")
```


## Step 1: Understand the relationship between horsepower and fuel efficiency
```{r}
summary(mtcars[, c("mpg", "hp")])
```


## Step 2: Create a scatter plot to visualize the relationship

### Principle 1: Clarity - Scatter plot showing how mpg (fuel efficiency) relates to hp (horsepower)
```{r}
library(ggplot2)
ggplot(mtcars, aes(x = hp, y = mpg, label = rownames(mtcars))) +
  geom_point(color = "blue", size = 3) +
  geom_text(vjust = -0.5, size = 3) +  # Add car names to the points
  theme_minimal() +
  labs(title = "Fuel Efficiency vs Horsepower", 
       x = "Horsepower (hp)", 
       y = "Miles per Gallon (mpg)") +
  theme(plot.title = element_text(hjust = 0.5))
```


## Step 3: Simple Analysis

### Principle 2: Simplicity - We filter cars with higher-than-average mpg and horsepower
```{r}

avg_mpg <- mean(mtcars$mpg)
avg_hp <- mean(mtcars$hp)

efficient_powerful_cars <- subset(mtcars, mpg > avg_mpg | hp > avg_hp)

cat("\nCars that provide a balance of good fuel efficiency and power:\n")
print(efficient_powerful_cars)
```


## Step 4: Visualize best cars

### Principle 3: Consistency - A consistent format is used to highlight selected cars
```{r}


ggplot(efficient_powerful_cars, aes(x = hp, y = mpg, label = rownames(efficient_powerful_cars))) +
  geom_point(color = "green", size = 4) +
  geom_text(vjust = -0.5, size = 3) +  # Add car names
  theme_minimal() +
  labs(title = "High Efficiency & Power Cars", 
       x = "Horsepower (hp)", 
       y = "Miles per Gallon (mpg)") +
  theme(plot.title = element_text(hjust = 0.5))

```

