---
title: "Untitled"
author: "Idriss"
date: "`r Sys.Date()`"
output: html_document
---

# PORTFOLIO FINANCIAL ANALYSIS

In portfolio financial analysis, several models and techniques are commonly used to assess risk, return, and overall performance. Here are some key models:

### 1. **Modern Portfolio Theory (MPT)**
- **Developed by:** Harry Markowitz
- **Concept:** Investors can construct portfolios to maximize expected return based on a given level of market risk, emphasizing diversification.
- **Key Metric:** Efficient Frontier, which represents the set of optimal portfolios offering the highest expected return for a defined level of risk.

### 2. **Capital Asset Pricing Model (CAPM)**
- **Concept:** Describes the relationship between systematic risk and expected return for assets, particularly stocks.
- **Formula:** 
  \[ E(R_i) = R_f + \beta_i (E(R_m) - R_f) \]
  Where \(E(R_i)\) is the expected return on the capital asset, \(R_f\) is the risk-free rate, \(\beta_i\) is the beta of the security, and \(E(R_m)\) is the expected market return.

### 3. **Arbitrage Pricing Theory (APT)**
- **Concept:** An alternative to CAPM, APT considers multiple factors that might affect an asset's return, instead of just market risk.
- **Formula:** 
  \[ E(R_i) = R_f + \beta_{i1}F_1 + \beta_{i2}F_2 + ... + \beta_{in}F_n \]
  Where \(F\) represents various factors.

### 4. **Fama-French Three-Factor Model**
- **Developed by:** Eugene Fama and Kenneth French
- **Concept:** Expands on CAPM by adding size risk (SMB, Small Minus Big) and value risk (HML, High Minus Low) factors.
- **Formula:**
  \[ E(R_i) = R_f + \beta_{i1}(E(R_m) - R_f) + \beta_{i2}SMB + \beta_{i3}HML \]

### 5. **Black-Litterman Model**
- **Developed by:** Fischer Black and Robert Litterman
- **Concept:** Combines investor views with market equilibrium to estimate the expected returns of assets.
- **Use:** Often used in portfolio optimization to create a more stable and intuitive estimation of expected returns.

### 6. **Value at Risk (VaR)**
- **Concept:** Measures the potential loss in value of a portfolio over a defined period for a given confidence interval.
- **Usage:** Helps in risk management to understand the maximum potential loss.

### 7. **Monte Carlo Simulation**
- **Concept:** Uses repeated random sampling to simulate the performance of a portfolio under various conditions.
- **Usage:** Useful for estimating the impact of risk and uncertainty in financial and investment decisions.

### 8. **Sharpe Ratio**
- **Concept:** Measures the performance of an investment compared to a risk-free asset, after adjusting for its risk.
- **Formula:** 
  \[ \text{Sharpe Ratio} = \frac{E(R_p) - R_f}{\sigma_p} \]
  Where \(E(R_p)\) is the expected portfolio return, \(R_f\) is the risk-free rate, and \(\sigma_p\) is the portfolio standard deviation.

These models and tools are essential for constructing, analyzing, and managing portfolios to achieve desired risk-return profiles.

# getting the data

## Installing and importing thelibrary
```{r}
install.packages("quantmod")
library(quantmod)
```


## importing data from yahoo

```{r}
# Load the quantmod package
library(quantmod)

# Get S&P 500 (SPY) data
getSymbols("SPY", src = "yahoo", from = "2000-01-01", to = Sys.Date())
spy_data <- SPY

# Get Dow Jones Industrial Average (DJI) data
getSymbols("^DJI", src = "yahoo", from = "2000-01-01", to = Sys.Date())
dji_data <- DJI

```

